% Make sure to keep the text within the brackets and the output should turn out correct
\acknowledge{%
  At the outset, I would like to thank my advisor Dr. Beverly Sanders for providing
  me the opportunity to work on this project and all the guidance throughout the
  course of my thesis work. She provided new ideas whenever I got stuck
  and helped me with finer points. And most importantly she
  introduced me to a whole new area of computation which I wouldn't have known
  to exist if I hadn't got a chance to work on this project.\\

  I would like to thank Dr. Ajith Perera for helping me in the matters of computational
  chemistry. As a domain scientist, he helped me in understanding domain problems
  and benchmarking my experiments on real life jobs.\\

  I would like to thank my committee members Dr. Erik Deumens and Dr. Alin Dobra
  for their advice and feedback on my thesis and research.\\

  I am deeply grateful to my family for the constant support and my brother for
  encouraging me to pursue higher education at a foreign university. And, I would
  like to thank my friend, Snehal, for keeping me motivated and helping me in
  writing my thesis.\\

  The author acknowledges University of Florida Research Computing for providing
  computational resources and support that have contributed to the research results
  reported in this thesis.​\\

  This material is based on research sponsored by AFRL/RXAP under grant FA8650-17-1-5401.
  The U.S Government is authorized to reproduce and distribute reprints for Governmental
  purposes notwithstanding any copyright notation thereon. The views and conclusions
  contained herein are those of the authors and should not be interpreted as
  necessarily representing the official policies or endorsements, either expressed
  or implied, of the AFRL/RXAP or the U.S. Government.
  }
